---
title: "EDA"
output: html_document
date: "2024-07-07"
---

# Setting up

This first chunk installs necessary packages. In the next code chunk, you will find the command to change the working directory. After you clone your repo, you will need to run this command ONCE. Do not push this change! Better to do it on the console and not change this code block

```{r}
# This command sets the wd for the R project
```


```{r}
#setwd("/Users/h.abdulla/Dropbox/Mac (3)/Desktop/Imperial/Modules/Y2/Term 3/Healthcare Analytics/Group Assignment/Healthcare Project")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("haven")
```

```{r}
library(haven)
library(tidyr)
library(ggplot2)
library(dplyr)
library(stargazer)
library(stringr)
library(naniar)
```

```{r}
# Change this if you saves the file in a different path
file = "gss_spss_with_codebook/GSS7218_R3.sav"

# Read the data file
raw_data <- read_sav(file)
```


```{r}
data_2016 <- raw_data %>%
  filter(YEAR == 2016)

# Calculate the total number of rows in the dataframe
total_rows <- nrow(data_2016)

# Calculate the percentage of non-NA values for each column
non_na_percentage <- colSums(!is.na(data_2016)) / total_rows

# Identify columns with more than 10% non-NA values
columns_with_more_than_10_percent_non_na <- names(non_na_percentage[non_na_percentage > 0.40])

# Print the column names
print(columns_with_more_than_10_percent_non_na)
```

# Data cleaning

By inspecting the codebook, we selected the following variables to take a closed look at (i.e: Our control variables)

```{r}
# Selecting candidate variables

sliced_data <- raw_data %>%
  select(
    # First, start with Mental health related variables
    # These are the Health variables
    SNSMOTH1   # DO YOU USE ANY OTHER SOCIAL NETWORKS
    #, SNSMOT2A #  FIRST OTHER SOCIAL NETWORK
    #, SNSMOT2B #  SECOND OTHER SOCIAL NETWORK
    #, SNSMOT2C #  THIRD OTHER SOCIAL NETWORK
   # , SNSMNEW  # WHICH SOCIAL NETWORK DID YOU JOIN MOST RECENTLY
   # , SNSMYEAR #  YEAR FIRST JOINED SNS
   # , SNSMFRST #  WHICH SOCIAL NETWORK DID YOU JOIN FIRST
    , INTWKDYM #  MINUTES OF INTERNET USE ON WEEKDAYS
    , INTWKDYH #  HOURS OF INTERNET USE ON WEEKDAYS
    , INTWKENM #  MINUTES OF INTERNET USE ON WEEKENDS
    , INTWKENH #  HOURS OF INTERNET USE ON WEEKENDS
    #, NUMPROBS #  HOW MANY FRIENDS CLOSE TO DISCUSS PROBLEMS
    #, BYEMPROB #  HOW MANY FRIENDS DISCUSS PROBLEMS VIA EMAIL
   # , INTACCSS #  DOES R HAVE INTERNET THROUGH CELL PHONE OR PDA
    , MNTLHLTH #  DAYS OF POOR MENTAL HEALTH PAST 30 DAYS
    #, WEIGHT   # Weight
    #, HEIGHT   # R IS HOW TALL
    #, LIFENOW  # R'S RATING OF LIFE OVERALL NOW FROM 0-10
    #, DEPRESS  # TOLD HAVE DEPRESSION
    #, HLTHPHYS # R'S PHYSICAL HEALTH
    #, QUALLIFE # R'S QUALITY OF LIFE
    #, HLTHMNTL # R'S MENTAL HEALTH, MOOD, AND ABILITY TO THINK.        
    , TWITTER  # CREATE DUMMY VARIABLE FOR THESE. IF ANY IS TRUE, MAKE IT 1, ELSE 0
    , LINKEDIN
    , SNAPCHAT
    , INSTAGRM
    , FACEBOOK
    
    # Then, control variables
    , YEAR     # GSS YEAR FOR THIS RESPONDENT
    
      # Labour
    , ID       # RESPONDENT ID NUMBER
    , WRKSTAT  # LABOR FORCE STATUS
    , HRS1     # NUMBER OF HOURS WORKED LAST WEEK
   # , HRS2     # NUMBER OF HOURS USUALLY WORK A WEEK
    #, COMMUTE  # TRAVEL TIME TO WORK
    , 
    , # Social
    , SEX      # RESPONDENTS SEX
    , RACE     # RACE OF RESPONDENT
    , BORN     # WAS R BORN IN THIS COUNTRY
    , REGION   # REGION OF INTERVIEW
   # , INCOME06 # TOTAL FAMILY INCOME
    #, RINCOM06 # RESPONDENTS INCOME
    , INCOME16 # TOTAL FAMILY INCOME
    , RINCOM16 # RESPONDENTS INCOME
    , CHILDS   # NUMBER OF CHILDREN
    , PAEDUC   # HIGHEST YEAR SCHOOL COMPLETED, FATHER
    , MAEDUC   # HIGHEST YEAR SCHOOL COMPLETED, MOTHER
    , SEXSEX   # Sex of sex partners in the last year
    , PRES16   # How voted for (i.e: liberal or not)
    , 
    
    # New Variables to consider
    , HOMPOP #NUMBER OF PERSONS IN HOUSEHOLD
    , EDUC #HIGHEST YEAR OF SCHOOL COMPLETED                                      
    , SPWRKSTA #SPOUSE LABOR FORCE STATUS
    , MARITAL #MARITAL STATUS
  )

```


Since we are interested in the impact of social media usage on people's metal health (perceived), we will limit the scope of the data to the responses from 2006 and 2016

```{r}

# Plot bar plot for 'Exercise' variable
ggplot(raw_data, aes(x = YEAR)) +
  geom_bar() +
  labs(title = "Number of reponses by year", x = "Year", y = "Count") +
  theme_minimal()

```

We see that it has a similar level of responses to the other years so there are no concerns with regards to the availability of the data.

```{r}
filtered_data <- sliced_data %>%
  filter(YEAR %in% c(2006, 2016))
```

Next let's check for missing variables

```{r}
vis_miss(filtered_data)
```

We assume the variables where 100% of the data is missing to be not present in the latest version of the survey. Therefore we remove them from the target variables. We also removed the "Number of hours worked last week" because there is already a similar question in the codebook and for this variable 98% of the data is missing. Same was done for SNSMOT2A:

```{r}
sliced_data <- filtered_data %>%
  select(
     # First, start with Mental health related variables
    # These are the Health variables
    SNSMOTH1   # DO YOU USE ANY OTHER SOCIAL NETWORKS
    #, SNSMOT2A #  FIRST OTHER SOCIAL NETWORK
    #, SNSMOT2B #  SECOND OTHER SOCIAL NETWORK
    #, SNSMOT2C #  THIRD OTHER SOCIAL NETWORK
   # , SNSMNEW  # WHICH SOCIAL NETWORK DID YOU JOIN MOST RECENTLY
   # , SNSMYEAR #  YEAR FIRST JOINED SNS
   # , SNSMFRST #  WHICH SOCIAL NETWORK DID YOU JOIN FIRST
    , INTWKDYM #  MINUTES OF INTERNET USE ON WEEKDAYS
    , INTWKDYH #  HOURS OF INTERNET USE ON WEEKDAYS
    , INTWKENM #  MINUTES OF INTERNET USE ON WEEKENDS
    , INTWKENH #  HOURS OF INTERNET USE ON WEEKENDS
    #, NUMPROBS #  HOW MANY FRIENDS CLOSE TO DISCUSS PROBLEMS
    #, BYEMPROB #  HOW MANY FRIENDS DISCUSS PROBLEMS VIA EMAIL
   # , INTACCSS #  DOES R HAVE INTERNET THROUGH CELL PHONE OR PDA
    , MNTLHLTH #  DAYS OF POOR MENTAL HEALTH PAST 30 DAYS
    #, WEIGHT   # Weight
    #, HEIGHT   # R IS HOW TALL
    #, LIFENOW  # R'S RATING OF LIFE OVERALL NOW FROM 0-10
    #, DEPRESS  # TOLD HAVE DEPRESSION
    #, HLTHPHYS # R'S PHYSICAL HEALTH
    #, QUALLIFE # R'S QUALITY OF LIFE
    #, HLTHMNTL # R'S MENTAL HEALTH, MOOD, AND ABILITY TO THINK.        
    , TWITTER  # CREATE DUMMY VARIABLE FOR THESE. IF ANY IS TRUE, MAKE IT 1, ELSE 0
    , LINKEDIN
    , SNAPCHAT
    , INSTAGRM
    , FACEBOOK
    
    # Then, control variables
    , YEAR     # GSS YEAR FOR THIS RESPONDENT
    
      # Labour
    , ID       # RESPONDENT ID NUMBER
    , WRKSTAT  # LABOR FORCE STATUS
    , HRS1     # NUMBER OF HOURS WORKED LAST WEEK
   # , HRS2     # NUMBER OF HOURS USUALLY WORK A WEEK
    #, COMMUTE  # TRAVEL TIME TO WORK
    , 
    , # Social
    , SEX      # RESPONDENTS SEX
    , RACE     # RACE OF RESPONDENT
    , BORN     # WAS R BORN IN THIS COUNTRY
    , REGION   # REGION OF INTERVIEW
   # , INCOME06 # TOTAL FAMILY INCOME
    #, RINCOM06 # RESPONDENTS INCOME
    , INCOME16 # TOTAL FAMILY INCOME
    , RINCOM16 # RESPONDENTS INCOME
    , CHILDS   # NUMBER OF CHILDREN
    , PAEDUC   # HIGHEST YEAR SCHOOL COMPLETED, FATHER
    , MAEDUC   # HIGHEST YEAR SCHOOL COMPLETED, MOTHER
    , SEXSEX   # Sex of sex partners in the last year
    , PRES16   # How voted for (i.e: liberal or not)
    , 
    
    # New Variables to consider
    , HOMPOP #NUMBER OF PERSONS IN HOUSEHOLD
    , EDUC #HIGHEST YEAR OF SCHOOL COMPLETED                                      
    , SPWRKSTA #SPOUSE LABOR FORCE STATUS
    , MARITAL #MARITAL STATUS
  )
  
```

```{r}
vis_miss(sliced_data)
```

Our selected dependent variable has only 62% of the data present. We try to investigate further it there is some pattern to the missing data. We immediately observe that the questions about social media usage were not present prior to 2016, as their missing values match perfectly with the ones to the income question. We can validate this assumtion by counting the non-na values in 2016. Therefore, to use this variables we will need to keep the timeframe restricted to 2016

```{r}
# Filtering the data again
filtered_data <- sliced_data %>%
  filter(YEAR == 2016)

#Slicing the data again
sliced_data <- filtered_data %>%
  select(
     # First, start with Mental health related variables
    # These are the Health variables
    SNSMOTH1   # DO YOU USE ANY OTHER SOCIAL NETWORKS
    #, SNSMOT2A #  FIRST OTHER SOCIAL NETWORK
    #, SNSMOT2B #  SECOND OTHER SOCIAL NETWORK
    #, SNSMOT2C #  THIRD OTHER SOCIAL NETWORK
   # , SNSMNEW  # WHICH SOCIAL NETWORK DID YOU JOIN MOST RECENTLY
   # , SNSMYEAR #  YEAR FIRST JOINED SNS
   # , SNSMFRST #  WHICH SOCIAL NETWORK DID YOU JOIN FIRST
    , INTWKDYM #  MINUTES OF INTERNET USE ON WEEKDAYS
    , INTWKDYH #  HOURS OF INTERNET USE ON WEEKDAYS
    , INTWKENM #  MINUTES OF INTERNET USE ON WEEKENDS
    , INTWKENH #  HOURS OF INTERNET USE ON WEEKENDS
    #, NUMPROBS #  HOW MANY FRIENDS CLOSE TO DISCUSS PROBLEMS
    #, BYEMPROB #  HOW MANY FRIENDS DISCUSS PROBLEMS VIA EMAIL
   # , INTACCSS #  DOES R HAVE INTERNET THROUGH CELL PHONE OR PDA
    , MNTLHLTH #  DAYS OF POOR MENTAL HEALTH PAST 30 DAYS
    #, WEIGHT   # Weight
    #, HEIGHT   # R IS HOW TALL
    #, LIFENOW  # R'S RATING OF LIFE OVERALL NOW FROM 0-10
    #, DEPRESS  # TOLD HAVE DEPRESSION
    #, HLTHPHYS # R'S PHYSICAL HEALTH
    #, QUALLIFE # R'S QUALITY OF LIFE
    #, HLTHMNTL # R'S MENTAL HEALTH, MOOD, AND ABILITY TO THINK.        
    , TWITTER  # CREATE DUMMY VARIABLE FOR THESE. IF ANY IS TRUE, MAKE IT 1, ELSE 0
    , LINKEDIN
    , SNAPCHAT
    , INSTAGRM
    , FACEBOOK
    
    # Then, control variables
    , YEAR     # GSS YEAR FOR THIS RESPONDENT
    
      # Labour
    , ID       # RESPONDENT ID NUMBER
    , WRKSTAT  # LABOR FORCE STATUS
    , HRS1     # NUMBER OF HOURS WORKED LAST WEEK
   # , HRS2     # NUMBER OF HOURS USUALLY WORK A WEEK
    #, COMMUTE  # TRAVEL TIME TO WORK
    , 
    , # Social
    , SEX      # RESPONDENTS SEX
    , RACE     # RACE OF RESPONDENT
    , BORN     # WAS R BORN IN THIS COUNTRY
    , REGION   # REGION OF INTERVIEW
   # , INCOME06 # TOTAL FAMILY INCOME
    #, RINCOM06 # RESPONDENTS INCOME
    , INCOME16 # TOTAL FAMILY INCOME
    , RINCOM16 # RESPONDENTS INCOME
    , CHILDS   # NUMBER OF CHILDREN
    , PAEDUC   # HIGHEST YEAR SCHOOL COMPLETED, FATHER
    , MAEDUC   # HIGHEST YEAR SCHOOL COMPLETED, MOTHER
    , SEXSEX   # Sex of sex partners in the last year
    , PRES16   # How voted for (i.e: liberal or not)
    , 
    
    # New Variables to consider
    , HOMPOP #NUMBER OF PERSONS IN HOUSEHOLD
    , EDUC #HIGHEST YEAR OF SCHOOL COMPLETED                                      
    , SPWRKSTA #SPOUSE LABOR FORCE STATUS
    , MARITAL #MARITAL STATUS
  )
```


```{r}
gg_miss_upset(sliced_data, nsets = 10, nintersects = 50)
```
By inspecting the upset vizualition, we could not find any significant pattern with the dependent variable. Therefore, we will simply drop the na's from consideration

```{r}
# Drop NA's and the no answers: IAP, DONT KNOW and NO ANSWER values
filtered_data <- filtered_data %>%
  filter(!is.na(MNTLHLTH) & !MNTLHLTH %in% c(-1, 98, 99))

ggplot(filtered_data, aes(x = MNTLHLTH)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(title = "Distribution of days with poor mental health", x = "Number of days with poor mental health", y = "Count") +
  theme_minimal()
```


# EDA

We start our exploration by looking at the amount of minutes spent on internet. We want to understand if that is a complement to the hoursof internet usage. We start by plotting the histogram:

```{r}

#    , INTWKDYM #  MINUTES OF INTERNET USE ON WEEKDAYS
#    , INTWKDYH #  HOURS OF INTERNET USE ON WEEKDAYS
#    , INTWKENM #  MINUTES OF INTERNET USE ON WEEKENDS
#    , INTWKENH #  HOURS OF INTERNET USE ON WEEKENDS

ggplot(filtered_data, aes(x = INTWKDYM)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(title = "Distribution of minutes of internet usage per weekday", x = "Minutes", y = "Count") +
  theme_minimal()
```
Same for weekends

```{r}

#    , INTWKDYM #  MINUTES OF INTERNET USE ON WEEKDAYS
#    , INTWKDYH #  HOURS OF INTERNET USE ON WEEKDAYS
#    , INTWKENM #  MINUTES OF INTERNET USE ON WEEKENDS
#    , INTWKENH #  HOURS OF INTERNET USE ON WEEKENDS

ggplot(filtered_data, aes(x = INTWKENM)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(title = "Distribution of minutes of internet usage per weekday", x = "Minutes", y = "Count") +
  theme_minimal()
```
As the range is always capped at 60, our assumption is that this is either a complement to the number of hours or applicable when respondents do not spend even 1 hour using the internet per week. For simplicity of analysis and since we are interested in the relantionship between the total amount of time spent per week on social media, we will create a new variable that returns 1 if the number of minutes spent is >= 20 minutes and simply round to zero otherwise. This will also take care of the duplicates

```{r}
data <- filtered_data

# Create new column based on minutes per week
data$hours_wkd_from_m <- ifelse(data$INTWKDYM < 20, 0, 1)

# Create new column based on minutes per weekend
data$hours_wke_from_m <- ifelse(data$INTWKENM < 20, 0, 1)

```

Next, we look at the number of hours spend per week 

```{r}
ggplot(filtered_data, aes(x = INTWKDYH)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(title = "Distribution of minutes of internet usage per weekday", x = "Hours", y = "Count") +
  theme_minimal()

```


```{r}
ggplot(filtered_data, aes(x = INTWKENH)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(title = "Distribution of minutes of internet usage per weekends", x = "Hours", y = "Count") +
  theme_minimal()
```

In order to simplify the model, we will create a new variable that combines the total time spent on social media

```{r}
# Total time spent on the internet per week
data$total_internet_time <- data$hours_wkd_from_m + data$hours_wke_from_m + data$INTWKENH + data$INTWKDYH

# Plot total internet usage time per week
ggplot(data, aes(x = total_internet_time)) +
  geom_histogram(fill = "skyblue", color = "black") +
  labs(title = "Distribution of minutes of internet usage per weekends", x = "Hours", y = "Count") +
  theme_minimal()
```

```{r}
#    , SNSMNEW  # WHICH SOCIAL NETWORK DID YOU JOIN MOST RECENTLY

# Create a bar plot using ggplot2
ggplot(data, aes(x = SNSMYEAR)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Count of Categories", x = "Category", y = "Count") +
  theme_minimal()
```
Next, we plot the count of RACE responses from the survey

```{r}
# Create a bar plot using ggplot2
ggplot(data, aes(x = RACE)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Count of Categories - Race", x = "Category", y = "Count") +
  theme_minimal()
```
We see that approximately 80% of the respondents are white. Therefore, we will create an auxiliary dummy variable for "is white" and use it as an iteration term

```{r}

# Creating dummy variable for "white"
data$flg_is_white <- ifelse(data$RACE == 1, 1, 0)

```


# Modeling

```{r}
data$flg_is_male <- ifelse(data$SEX == 1,1,0)
```


```{r}
ols1 = lm(
  MNTLHLTH  
  ~ BORN
  + REGION
  + WRKSTAT  # LABOR FORCE STATUS
  + HRS1     # NUMBER OF HOURS WORKED LAST WEEK
  + INCOME16 # TOTAL FAMILY INCOME
  + RINCOM16 # Respondant's income
  + SNSMOTH1 # DO YOU USE ANY OTHER SOCIAL NETWORKS
  + total_internet_time*flg_is_male*flg_is_white
  + SNSMYEAR # SNSMYEAR
  , data=data)

stargazer(ols1,
          type="text",
          single.row = TRUE,
          header = FALSE,
          title = "Simple regression - No changes")
```


```{r}
ols1 = lm(
  MNTLHLTH
  ~ BORN
  + REGION
  + WRKSTAT  # LABOR FORCE STATUS
  + HRS1     # NUMBER OF HOURS WORKED LAST WEEK
  + INCOME16 # TOTAL FAMILY INCOME
  + RINCOM16 # Respondant's income
  + SNSMOTH1 # DO YOU USE ANY OTHER SOCIAL NETWORKS
  + total_internet_time*flg_is_male*flg_is_white
  + SNSMYEAR # SNSMYEAR
  + HOMPOP
  + HEALTH1
  , data=data)

stargazer(ols1,
          type="text",
          single.row = TRUE,
          header = FALSE,
          title = "Simple regression - No changes")
```


